Towards Patient-Specific Neural Field Modeling using Micro-Electrode Data

### 2011-02-06 ltx/data_paper/DataPaper.tex
edits to E step

Since the state equation is non-linear,  calculation of covariance and  cross-covariance terms required for the maximisation step is not straightforward as in the linear case. Therefore, an additional element to the URTSS algorithm is also created to approximate $q(\mathbf{x}_t)$, which is required in the M-step. This is accomplished by taking the first-order truncated Taylor-series expansion of the sigmoidal activation function  around the smoothed state, $\hat{\mathbf x}_t^{b}$ at each time instant, i.e.
\begin{equation}\label{eq:SigmoidTaylor}
 f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\mathbf x_t\right) \approx f\left(\boldsymbol \phi^\top(\mathbf r')\hat{\mathbf x}_t^{b}\right)+  \boldsymbol \phi^\top\left(\mathbf r'\right)\left(\mathbf x_t - \hat{\mathbf x}_t^{b}\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^{b}\right)
\end{equation}
where 
\begin{equation}\label{eq:SigmoidDerivative}
 f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^{b}\right)=\varsigma f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^{b}\right)\left( 1-f\left( \boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^{b}\right)\right)
\end{equation}
substituting \ref{eq:SigmoidTaylor} in \ref{eq:QmatrixForSigmapoints}  we get an approximation for $q(\mathbf x_t)$ as
\begin{equation}\label{eq:qTaylor}
 q(\mathbf x_t) \approx q\left(\hat{\mathbf x}_t^{b}\right)+\int_\Omega \boldsymbol{\Psi}\left(\mathbf{r}'\right)\boldsymbol \phi^\top\left(\mathbf r'\right) \left(\mathbf x_t -\hat{\mathbf x}_t^{b}\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^{b}\right) d\mathbf{r}'
\end{equation}
Equation \ref{eq:qTaylor} is used to approximate the lower bound on the log-likelihood function in the M-step.

@@

Substituting \ref{eq:qTaylor} in \ref{eq:Qfunctionintrace} and taking the expectation of the log likelihood function for all time instants gives an approximate to the lower-bound that needs to be maximised to gain optimal parameter estimates. Therefore $\mathcal Q$-function can be approximated as  

>>>>

Since the state equation is non-linear,  calculation of covariance and  cross-covariance terms required for the maximisation step is not straightforward as in the linear case. Therefore, an additional  Taylor expansion step to the URTSS algorithm is also created to approximate $q(\mathbf{x}_t)$, which is required in the M-step. This is accomplished by taking the first-order truncated Taylor-series expansion of the sigmoidal activation function  around the smoothed state, $\hat{\mathbf x}_t^{b}$ at each time instant, i.e.
\begin{equation}\label{eq:SigmoidTaylor}
 f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\mathbf x_t\right) \approx f\left(\boldsymbol \phi^\top(\mathbf r')\hat{\mathbf x}_t^{b}\right)+  \boldsymbol \phi^\top\left(\mathbf r'\right)\left(\mathbf x_t - \hat{\mathbf x}_t^{b}\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^{b}\right)
\end{equation}
where 
\begin{equation}\label{eq:SigmoidDerivative}
 f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^{b}\right)=\varsigma f\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^{b}\right)\left( 1-f\left( \boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^{b}\right)\right)
\end{equation}
substituting \ref{eq:SigmoidTaylor} in \ref{eq:QmatrixForSigmapoints}  we get an approximation for $q(\mathbf x_t)$ as
\begin{equation}\label{eq:qTaylor}
 q(\mathbf x_t) \approx q\left(\hat{\mathbf x}_t^{b}\right)+\int_\Omega \boldsymbol{\Psi}\left(\mathbf{r}'\right)\boldsymbol \phi^\top\left(\mathbf r'\right) \left(\mathbf x_t -\hat{\mathbf x}_t^{b}\right)f'\left(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^{b}\right) d\mathbf{r}'
\end{equation}
Equation \ref{eq:qTaylor} is used to approximate the lower bound on the log-likelihood function in the M-step.

@@

Having approximated $q\left(\mathbf x_t\right)$ in the E-step, Substituting \ref{eq:qTaylor} in \ref{eq:Qfunctionintrace} and taking the expectation of the log likelihood function for all time instants gives an approximate to the lower-bound that needs to be maximised to gain optimal parameter estimates. Therefore $\mathcal Q$-function can be approximated as  
